---
title: "Stop motion editing with stopmotion: the laser dinosaur"
vignette: >
  %\VignetteIndexEntry{Stop motion editing with stopmotion: the laser dinosaur}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
knitr:
  opts_chunk:
    collapse: true
    warning: false
    comment: "#>"
    fig.align: center
---

## Credit

The dinosaur animation used throughout this vignette was created by
**\@looksrawr** and is released under the
[CC0 1.0 Universal Public Domain Dedication](https://creativecommons.org/publicdomain/zero/1.0/).

---

## Overview

**stopmotion** is a pipeline-friendly toolkit for assembling and editing stop
motion animations from sequences of still images.  Its functions fall into
three families:

| Family | Functions |
|---|---|
| Load | `read()` |
| Restructure | `duplicate()`, `splice()` |
| Transform | `rotate()`, `wiggle()`, `flip()`, `flop()`, `blur()`, `scale()`, `crop()`, `trim()`, `border()`, `background()`, `centre()` |
| Display | `montage()`, `preview()` |

All functions accept an optional `frames` argument so that any operation can
target a precise subset of frames — the feature that sets **stopmotion** apart
from plain **magick** pipelines.

After each operation **stopmotion** prints a message listing the updated frame
sequence.  These messages are shown in interactive sessions and suppressed
automatically during document rendering.  To control verbosity explicitly, use
`stopmotion_verbosity()` or set the option directly:

```{r options}
#| eval: true
library(stopmotion)

stopmotion_verbosity(FALSE) # silence for the rest of the session
options(stopmotion.verbose = FALSE) # equivalent
```
Or edit your `.Rprofile` to save the option with `usethis::edit_r_profile()`

This vignette walks through a complete editing session using the bundled
`extdata/` frames: a ten-frame cartoon dinosaur whose eyes shoot a laser ray.

---

## 1  Loading a GIF

The bundled `extdata/` directory contains the ten frames of the dinosaur
animation as individual PNG files — exactly the kind of image sequence that
**stopmotion** is designed to work with.  `read()` loads them in lexicographic
order and returns a `magick-image` object accepted by every **stopmotion**
function.

```{r load}
library(magick)

dino_dir <- system.file("extdata", package = "stopmotion")
dino <- read(dir = dino_dir)
dino
```

```{r frame-count}
cat("Number of frames:", length(dino), "\n")
image_info(dino)[, c("width", "height", "filesize")]
```

Ten frames, 480 × 480 pixels.

---

## 2  Inspecting frames

`montage` is a quick way to display all frames side-by-side.

```{r montage}
#| fig-width: 7
#| fig-height: 2.5
montage(dino, tile = "10x1", geometry = "64x64+2+2")
```

Scanning left to right you can read the story:

* **Frames 1–4** — the dinosaur stands still, then opens its eyes.
* **Frames 5–6** — eyes glow and start charging up.
* **Frames 7–9** — the laser fires.
* **Frame 10**   — the beam fades and the dinosaur looks pleased.

---

## 3  Building anticipation with `duplicate()`

The charging phase (frames 5–6) flashes by too quickly.  `duplicate()` with
`style = "looped"` inserts a copy of those frames immediately after the
originals, making the build-up feel more deliberate.

```{r duplicate}
dino2 <- duplicate(dino, frames = 5:6, style = "looped")
cat("Frames after duplicate():", length(dino2), "\n")
```

The `looped` style repeats the selected range in order, so the sequence
becomes …5, 6, 5, 6… — a natural charging pulse.

---

## 4  Adding drama with `border()`

A vivid red border on the laser frames (now frames 7–11 after the insertion)
signals "danger" to the viewer.

```{r border}
dino3 <- border(dino2, color = "red", geometry = "8x8", frames = 7:11)
```

---

## 5  Motion blur on the laser beam with `blur()`

The three peak laser frames (frames 8–10) benefit from a subtle blur that
conveys raw energy.

```{r blur}
dino4 <- blur(dino3, radius = 3, sigma = 1.5, frames = 8:10)
```

---

## 6  Hand-held shake with `wiggle()`

`wiggle()` inserts two slightly rotated copies after each selected frame —
one tilted `+degrees`, one `−degrees` — creating the organic wobble typical
of real stop motion.  We apply it to the quieter frames at the start so the
calm contrast with the explosive laser section.

```{r wiggle}
dino5 <- wiggle(dino4, degrees = 2, frames = 1:3)
cat("Total frames after wiggle():", length(dino5), "\n")
```

---

## 7  Putting it together: the full pipeline

The edits above can be chained into a single pipe for clarity.

```{r pipeline}
#| eval: true
read(dir = system.file("extdata", package = "stopmotion")) |>
  wiggle(degrees = 2, frames = 1:3) |>             # hand-held shake
  duplicate(frames = 5:6, style = "looped") |>     # hold the charge
  border(color = "red", geometry = "8x8",
          frames = 7:11) |>                        # danger border
  blur(radius = 3, sigma = 1.5, frames = 8:10) |>  # energy blur
  preview(fps = 2)
```

---

## 8  Exporting the result

`magick::image_write_gif()` writes the edited sequence back to a GIF file.
The `delay` argument controls playback speed (seconds per frame).

```{r export}
#| eval: false
out <- tempfile(fileext = ".gif")
image_write_gif(dino_final, path = out, delay = 1 / 8)
message("Saved to: ", out)
```

---

## 9  Other useful operations

### 9.1  Dropping unwanted frames with `splice()`

`splice()` inserts new frames after a given position.  Combined with standard
**magick** subsetting you can also remove frames:

```{r splice}
#| eval: false
# Insert a custom "RAWR!" title card after frame 4
title_card <- image_blank(480, 480, color = "black") |>
  image_annotate("RAWR!", size = 80, color = "red", gravity = "Center")

dino_with_title <- splice(dino, insert = title_card, after = 4L)
```

### 9.2  Scaling down for the web

`scale()` accepts any **magick** geometry string.

```{r scale}
#| eval: false
dino_small <- scale(dino, geometry = "50%")
```

### 9.3  Cropping to the face

```{r crop}
#| eval: false
# Keep a 200×200 window centred on the head (adjust offsets to taste)
dino_face <- crop(dino, geometry = "200x200+140+60")
```

### 9.4  Aligning frames with `centre()`

When frames drift slightly between photos, `centre()` performs an affine warp
guided by two reference points per frame (e.g. left and right eye positions)
so the subject stays locked in place across the whole animation.

```{r centre}
#| eval: false
# Record two stable landmarks per frame (e.g. with locator())
pts <- data.frame(
  frame = c(1L, 1L, 2L, 2L),
  x     = c(210, 270, 212, 269),
  y     = c(160, 162, 158, 161)
)
dino_stabilised <- centre(dino, points = pts, reference = 1L)
```

---

## Summary

| Step | Function | Key argument |
|---|---|---|
| Hold charging frames | `duplicate()` | `style = "looped"` |
| Red danger border | `border()` | `color`, `geometry` |
| Energy motion blur | `blur()` | `radius`, `sigma` |
| Hand-held shake | `wiggle()` | `degrees` |
| Insert title card | `splice()` | `insert`, `after` |
| Resize for web | `scale()` | `geometry` |
| Crop to face | `crop()` | `geometry` |
| Stabilise frames | `centre()` | `points`, `reference` |

All of the above accept a `frames` argument to restrict the operation to any
subset of frames, giving you frame-precise control over your animation.
